Sex-Specific Evolution of the Meiotic Recombination Rate

### 2020-06-13 doc/MS_v3.Rmd
first draft of abstract

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  html_document:
    toc: yes
    toc_depth: '5'
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '5'
  word_document:
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

@@

@@

<new>
We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 1). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus* and *M. spretus*. <removed _M. caroli_> We subsequently denote strains by their abbreviated subspecies and name (e.g. domesticusWSB). Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice from an inbred strain of Gough Island mice were sampled after **XX** <trck down the generation of Gough inbred levels> generations of brother-sister mating. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

<new>
The same dry-down spread technique was applied to both spermatocytes and oocytes, following (@peters_1997), with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used for MLH1 counts were between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of age were used for DMC1 counts. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20\*C if not stained immediately. To visualize the structure of meiotic chromosomes, we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)). Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37\*C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti-rabbit anti-DMC1 (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37\*C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37\* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS.

@@

<new 5/n>
We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We used mouse average as the dependent variable in all analyses. We first constructed a linear mixed model (M1) using lmer() from the lmer4 package [@lme4] in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT() from RLRsim [@RLRsim]). Subspecies, sex, and their interaction were coded as fixed effects, with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()) [@lme4]. As each observation was at the level of a single mouse (mouse average), it was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models. We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses using model M1, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFDraw, IFDnorm, and average MLH1 position (for single-focus bivalents). Four additional lnear models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from model M1.

@@

#adjust order of plot here
strain.table.HQ_Fig$strain <- factor(strain.table.HQ_Fig$strain,levels =c(
     "MSM", "MOLF",   
  "PWD", "SKIVE",  "KAZ",
  "WSB","G","LEW","PERC",
                     "CAST",
                         "SPRET","SPIC" ), order=T )
```

@@

@@

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the difference in counts between the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains (*musculus^PWD^* and _molossinus^MSM^_) exhibit higher counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23) of foci (Figure 1C). Strain mean MLH1 focus counts from females and males are uncorrelated (Spearman’s $\rho$ = 0.08; p = 0.84) across the set of strains.
**track down the correlation value**

@@

The effect of subspecies is no longer significant in a model treating all factors as fixed effects **(M2; p = 0.09)**, highlighting strain and sex as salient variables. Two strains exhibit particularly strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

@@

@@

That females have longer SCs is further supported by models that include covariates, which identify sex as the most consistently significant effect for total SC  (M1: p = `r Q1.lmer_totSC_M1_results$sex.results`; M2: p = `r sum.Q1.glm_totSC_M2$coefficients[,c(1,4)][9,2]`; M3: p =`r sum.Q1.glm_totSC_M3$coefficients[,c(1,4)][8,2]`) and 
short bivalent measures (M1: p = `r Q1.lmer_short_M1_results$sex.results`; M2: p = `r summ.Q1.glm_short.SC_M2$coefficients[,c(1,4)][8,2] / 10000`; M3: p = `r summ.Q1.glm_short.SC_M3$coefficients[,c(1,4)][8,2] / 10000`).

The existence of some subspecies and strain effects (M1 interaction: p = `r Q1.lmer_short_M1_results$interaction.pval`; M2: strain-SKIVE `r summ.Q1.glm_short.SC_M2$coefficients[7,c(1,4)][2]`,  SKIVE * male p = `r summ.Q1.glm_short.SC_M2$coefficients[13,c(1,4)][2]`) further indicates that SC length has evolved among strains and among subspecies. In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males have more MLH1 foci per cell. This pattern further suggests that in high-recombination strains, spermatocytes have less space in which to position additional crossovers than do oocytes.

@@

We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: `r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; t-test; p = `r Q1_1CO.pos_t.test.FULL$p.value` ). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = `r lmer.Q1.nrm1CO_M1_results$sex.results` ; M2: p = `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (all t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (t-test; p = `r Q1.IFD.PER_t.test_KAZ$p.value` ). Examination of IFD~norm~ distributions indicates that female values are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Analysis of models treating IFD~norm~ as the dependent variable supports the inference of stronger interference in males, with sex being the most significant variable (M1: p = `r Q1.IFD.PER_M1_lmer.results$sex.results` ; M2: p = `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`, M3: p = `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`).

In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers (Figure 4C) further suggests that interference distances vary more in females than in males and that males display a stronger telomeric bias in the placement of the distal crossover. In summary, controlling for differences in SC length (chromatin compaction) using IFD~norm~ indicates that interference is stronger in males, whereas consideration of IFD~raw~ shows that the sexes exhibit a similar level of interference on the physical scale.

@@

We used the contrast between males from high-recombination strains and males from low-recombination strains to identify features of the recombination landscape associated with evolutionary transitions in the genome-wide recombination rate. We considered proportions of bivalents with different numbers of crossovers, double-strand breaks, SC length, and crossover positions, including crossover interference. Ninety-six percent of single bivalents in our pooled dataset (n = `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")`) have either one or two MLH focus (Figure 3). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Figure 3).

High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` in _musculus^SKIVE^_, `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_ , and `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. _musculus^KAZ^_: p = `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = `r MSM.MOLF.2CO.prop.test$p.value`) and _musculus^PWD^_ vs. _molossinus^MSM^_: p = `r PWD.MSM.2CO.prop.test$p.value`. These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number on multiple bivalents.

To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of `r sum( (DMC1table.main$ncells.x) )`  early zygotene and `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from three low-recombination strains ( _musculus^KAZ^_, _domesticus^WSB^_ , and  _domesticus^G^_) and two high-recombination strains ( _musculus^PWD^_ and _molossinus^MSM^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test;  p < `r 1/1000000` ). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = `r ttest.HighLow.Z.pval` ). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision.

Although there is a positive correlation between total SC length and MLH1 foci when both traits are measured as mouse means (Spearman’s r = `r cor( mouse.total.skel$mean.skel, mouse.total.skel$mean.MLH1, method = c("spearman") )`; p = `r cor.test(mouse.total.skel$mean.skel, mouse.total.skel$mean.MLH1, method=c("spearman") )$p.value` ), total SC length only partially differentiates high-recombination strains from low-recombination strains (Figure 5). Whereas high-recombination strains as a group have significantly greater total SC length than low-recombination strains (t-test; p = `r Q2tot.sc_t.test_High.Low$p.value`), separate tests within subspecies show that the two strain categories differ within _M. m. molossinus_ (p = `r Q2tot.sc_t.test_High.Low_Mol$p.value` ) but not within _M. m. musculus_ (p = `r Q2tot.sc_t.test_High.Low_Musc$p.value`).

Additionally, mouse means for the reduced (short and long) bivalent datasets do not differ between high-recombination and low-recombination strains (t-test; short: p = `r Q2short.biv_t.test_High.Low$p.value` ; long: p = `r Q2long.biv_t.test_High.Low$p.value`). In a model with total SC length as the dependent variable, two subspecies effects are significant ( _M. m. musculus_ p = `r sum.Q2.glm_mus.av_totSC_M4[2,4]` , _M. m. molossinus_, p = `r sum.Q2.glm_mus.av_totSC_M4[3,][4]` ). In models with SC lengths of short and long bivalents as dependent variables, several subspecies and strain effects reach significance (p < 0.05), but they are not consistent across models. Collectively, these observations reveal that evolution of the genome-wide recombination rate is not strongly associated with evolution of SC length.

High-recombination and low-recombination strains do not differ in MLH1 focus positions on bivalents with a single focus, either for the full dataset (t-test; p = `r Q2.1COpos_Ttest$p.value` ) or within subspecies (t-test; _M. m. musculus_: p = `r Q2.1COpos_mus.av_Musc_ttest$p.value` ; _M. m. molossinus_ : p = `r Q2.1COpos_mus.av_Mol_ttest$p.value` ). While _domesticus^WSB^_ and _molossinus^MOLF^_ exhibit strain effects in a model with normalized crossover position as the dependent variable, these two strains do not differ in MLH1 count. These results show that evolution of crossover position on chromosomes with single crossovers is decoupled from evolution of the genome-wide recombination rate, at least at this scale of resolution.

High-recombination strains have greater inter-focal distances (in both raw and normalized measured) than low-recombination strains (t-test; IFD~norm~: p = `r t.test_mouse.av.IFD_PER$p.value` ; IFD~raw~: p = `r t.test_mouse.av.IFD_ABS$p.value`). This pattern holds within _M. m. musculus_ (t-test; IFD~norm~: p = `r  t.test_mouse.av.IFD_PER_MUSC$p.value`; IFD~raw~: p = `r  t.test_mouse.av.IFD_ABS_MUSC$p.value`), with a trend in the same direction within _M. m. molossinus_ (IFD~norm~: p= `r t.test_mouse.av.IFD_PER_MOL$p.value` ; IFD~raw~: p = `r t.test_mouse.av.IFD_ABS_MOL$p.value`).

>>>>

---
title: 'Sex-Specific Evolution of the Meiotic Recombination Rate'
date: '`r Sys.Date()`'
author: "April L. Peterson, Bret Payseur"
output:
  html_document:
  word_document:
  pdf_document:
    includes:
      in_header: header.tex
always_allow_html: yes
bibliography: refs_Main_Report.bib
csl: elife.csl
---

@@

# ABSTRACT

Although meiotic recombination is required for successful gametogenesis in most species that reproduce sexually, the rate of crossing over varies among individuals. Differences in recombination rate between females and males are perhaps the most striking form of this variation. Existing data fail to directly address the extent to which recombination experiences similar evolutionary pressures. To fill this gap, we measured meiotic recombination in both sexes for a panel of house mouse cross three subspecies. Using inbred strains and single cell immunohistochemistry allowed us to place sex-specific observations within the same genetic background and meiotic context. Our results indicate highly discordant evolutionary patterns in the two sexes. Whereas male recombination rates show evidence of rapid evolution over short evolutionary timescales, female recombination rates measured in the same strains are mostly static. These results strongly indicate that house mouse has two genome wide recombination rates which display distinct evolutionary trajectories.

@@

### 0.   Importance of recombination rate

### 1. Understanding levels of variation in gwRR

### 2.  The House Mouse is a great model for uncovering evolutionary patterns at a short timescale

### 3. What we accomplished in this paper

@@

<new>
We used a panel of wild-derived inbred strains of house mice (*Mus musculus*) and related murid species to profile natural genetic variation in recombination (Table 1). Our survey included 5 strains from *M. m. musculus*, 4 strains from *M. m. domesticus*, 2 strains from *M. m. molossinus*, 2 strains from *M. m. castaneus*, and 1 strain each from *M. spicilegus* and *M. spretus*. <removed _M. caroli_> We subsequently denote strains by their abbreviated subspecies and name (e.g. *domesticus^WSB^*). Mice were housed at dedicated, temperature-controlled facilities in the UW-Madison School of Medicine and Public Health, with the exception of mice from Gough Island, which were housed in a temperature-controlled facility in the UW-Madison School of Veterinary Medicine. Mice from an inbred strain of Gough Island mice were sampled after **XX** <track down the generation of Gough inbred levels> generations of brother-sister mating. All mice were provided with ad libitum food and water. Procedures followed protocols approved by IACUC.

@@

<new>
The same dry-down spread technique was applied to both spermatocytes and oocytes, following [@peters_1997], with adjustment for volumes. Spermatocyte spreads were collected and prepared as described in [@peterson2019]. The majority of mice used for MLH1 counts were between 5 and 12 weeks of age. Juvenile mice between 12 and 15 days of age were used for DMC1 counts. Both ovaries were collected from embryos (16-21 embryonic days) or neonates (0-48 hours after birth). Whole testes were incubated in 3ml of hypotonic solution for 45 minutes. Decapsulated ovaries were incubated in 300ul of hypotonic solution for 45 minutes. Fifteen microliters of cell slurry (masticated gonads) were transferred to 80ul of 2% PFA solution. Cells were fixed in this solution and dried in a humid chamber at room temperature overnight. The following morning, slides were treated with a Photoflow wash (Kodak, cite). Slides were stored at -20\*C if not stained immediately. To visualize the structure of meiotic chromosomes, we used antibody markers for the centromere (CREST) and lateral element of the synaptonemal complex (SC) (SYCP3). Crossovers (COs) were visualized as MLH1 foci. Double strand breaks (DSBs) were visualized as DMC1 foci. The staining protocol followed [@anderson1999] and [@koehler2002]. Antibody staining and slide blocking were performed in 1X antibody dilution buffer (ADB) (normal donkey serum (Jackson ImmunoResearch), 1X PBS, bovine serum albumin (Sigma), and Triton X-100 (Sigma)). Following a 30-minute blocking wash in ABD, each slide was incubated with 60ul of a primary antibody master mix for 48 hours at 37\*C. The master mix recipe contained polyclonal anti-rabbit anti-MLH1 (Calbiochem; diluted 1:50) or anti-rabbit anti-DMC1 (mix of DMC1), anti-goat polyclonal anti-SYCP3, (Abcam; diluted 1:50), and anti-human polyclonal antibody to CREST (Antibodies, Inc; diluted 1:200) suspended in ADB. Slides were washed twice in 50ml ADB before the first round of secondary antibody incubation for 12 hours at 37\*C. Alexa Fluor 488 donkey anti-rabbit IgG (Invitrgoen, location; diluted to 1:100) and Coumarin AMCA donkey anti-human IgG (Jackson ImmunoResearch; diluted to 1:200) were suspended in ADB. The last incubation of Alexa Fluor 568 donkey anti-goat (Invitrogen; diluted 1:100) was incubated at 1:100 for 2 hours at 37\* C. Slides were fixed with Prolong Gold Antifade (Invitrogen) for 24 hours after a final wash in 1x PBS.

@@

<new 5/n>
We used a series of statistical models to interpret patterns of variation in the recombination traits we measured (Table 2). We used mouse average as the dependent variable in all analyses. We first constructed a linear mixed model (M1) using lmer() from the lmer4 package [@lme4] in R (v3.5.2) [@Rstudio]. In this model, strain was coded as a random effect, with significance evaluated using a likelihood ratio test (using exactRLRT() from RLRsim [@RLRsim]). Subspecies, sex, and their interaction were coded as fixed effects, with significance evaluated using a chi-square test comparing the full and reduced models (drop1() and anova()) [@lme4]. As each observation was at the level of a single mouse (mouse average), it was uniquely coded within the dataset, nesting was implicit, each strain only occurs within one strain, and was not explicitly coded in our linear and mixed models. We used the subspecies effect to quantify divergence between subspecies and the (random) strain effect to quantify variation within subspecies in a sex-specific manner. In separate analyses using model M1, we considered mouse averages as dependent variables for each of the following traits: MLH1 count per cell, total SC length per cell, single bivalent SC length per cell, IFD~raw~, IFD~norm~, and average MLH1 position (for single-focus bivalents). Four additional lnear models containing only fixed effects (M2-M5) (Table 2) were used to further investigate results obtained from model M1.

@@

#adjust order of plot here
strain.table.HQ_Fig$strain <- factor(strain.table.HQ_Fig$strain,levels =c(
     "MSM", "MOLF",   
  "PWD", "SKIVE",  "KAZ",
  "WSB","G","LEW","PERC",
                     "CAST",
                         "SPRET","SPIC" ), order=T )

@@

sperman.corr.FM.pval <- cor.test( ( strain.table.HQ_Fig %>% filter(sex == "female") %>% filter(species == "M.musculus" ) )$mean_co, ( strain.table.HQ_Fig %>% filter(sex == "male") %>% filter(species == "M.musculus" ) )$mean_co, 
                        method = c("spearman") )

@@

Graphical comparisons reveal sex-specific dynamics to the evolution of genome-wide recombination rate (Figure 1A). First, MLH1 focus counts differ between females and males in most strains. Second, the difference in counts between the sexes varies among strains. Although most strains show more MLH1 foci in females, two strains (*musculus^PWD^* and _molossinus^MSM^_) exhibit higher counts in males. In females, numbers of MLH1 foci are evenly distributed around the sex-wide mean of approximately 25 (Figure 1B). In stark contrast, males largely separate into two groups of strains with high numbers (near 30) and low numbers (near 23) of foci (Figure 1C). Strain mean MLH1 focus counts from females and males are uncorrelated (Spearman’s $\rho$ = 0.08 - `r sperman.corr.FM`; p = 0.84 -- `r sperman.corr.FM.pval$p.value`) across the set of strains.

@@

The effect of subspecies is no longer significant in a model treating all factors as fixed effects **(M2; p = 0.09 -- )**, highlighting strain and sex as salient variables. Two strains exhibit particularly strong effects on MLH1 focus count (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][3]` ; *domesticus^LEW^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][4]`), with sex-strain interactions involving three strains (M3; *domesticus^G^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]`; *molossinus^MSM^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` ; *musculus^PWD^* p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`).

@@

In summary, two approaches for measuring SC length demonstrate that females have longer SCs (chromosome axes), even in strains in which males recombine more. This pattern implies that in high-recombination strains, spermatocytes have less space than oocytes in which to position additional crossovers.

@@

@@

<new 1/n>
We used normalized positions of MLH1 foci along bivalents with a single focus to compare crossover location while controlling for differences in SC length. In all strains, MLH1 foci tend to be closer to the telomere in males (mean normalized position in males: 0.68--`r mean( (mouse.level.BivData_table %>% filter(sex == "male") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; mean normalized position in females: 0.56 --  `r mean( (mouse.level.BivData_table %>% filter(sex == "female") %>% filter(species == "M.musculus") )$norm.first.foci.pos, na.rm = TRUE )`; t-test; p = 3.84^{-23} -- `r Q1_1CO.pos_t.test.FULL$p.value`). Sex is also the strongest determinant of MLH1 focus position in the models we tested (M1: p = 7.79^{-26 --  `r lmer.Q1.nrm1CO_M1_results$sex.results`}; M2: p = 5.39^{-8} --  `r summry.Q1.glm_F1pos_M2$coefficients[,4][8]`; M3: p = 5.39^{-8} --  `r summry.Q1.glm_1CO.pos_M3$coefficients[,4][8]`).

<new 2/n>
Males have longer normalized mean inter-focal distances (IFD~norm~) than females in seven out of eight strains (t-tests; p < 0.05), with only _musculus^KAZ^_ showing no difference (p = 0.33 -- `r Q1.IFD.PER_t.test_KAZ$p.value`). Examination of IFD~norm~ distributions indicates that females are centered at approximately 50% and show a slight enrichment of low (<25%) values, whereas males are enriched for higher values. Models treating IFDnorm as the dependent variable support the inference of stronger interference in males, with sex being the most significant variable (M1: p = 3.11^{-12} -- `r Q1.IFD.PER_M1_lmer.results$sex.results` ; M2: p = 0.01  --  `r summry.glm.Q1_IFD.PER.M2$coefficients[,4][4]`; M3: p = 0.01  --  `r summry.glm.Q1_IFD.PER.M3$coefficients[,4][2]`). In contrast, there is no clear signal of sex differences in raw mean inter-focal distances (IFD~raw~) across the full set of strains, whether they are considered separately or together. Visualization of normalized MLH1 foci positions on bivalents with two crossovers (Figure 4C) further suggests that interference distances vary more in females than in males, and that males display a stronger telomeric bias in the placement of the distal crossover.

In summary, controlling for differences in SC length (chromatin compaction) indicates that interference is stronger in males, whereas interference on the physical scale is similar in the sexes.

@@

<new 1/n>
We used the contrast between males from high-recombination strains and males from low-recombination strains to identify features of the recombination landscape associated with evolutionary transitions in the genome-wide recombination rate. We considered proportions of bivalents with different numbers of crossovers, double-strand break number, SC length, and crossover positioning.

<new 2/n>
Ninety-six percent of single bivalents in our pooled dataset (n = 9,576 -- `r formatC( sum( Chrm.prop_strain_table$n0CO, Chrm.prop_strain_table$n1CO, Chrm.prop_strain_table$n2CO, Chrm.prop_strain_table$n3CO, Chrm.prop_strain_table$n4CO ), format="d", big.mark=",")` ) have either one or two MLH foci (Figure 3). The proportions of single-focus (1CO) bivalents vs. double-focus (2CO) bivalents distinguish high-recombination strains from low-recombination strains (Figure 3). High-recombination strains are enriched for 2CO bivalents at the expense of 1CO bivalents: proportions of 2CO bivalents are 0.33 -- `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "SKIVE"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "SKIVE"]` in _musculus^SKIVE^_, 0.44 -- `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "PWD"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "PWD"]` in _musculus^PWD^_ , and 0.51 -- `r chrm.prop.male.high1$n2CO[chrm.prop.male.high1$strain == "MSM"] / chrm.prop.male.high1$nbiv[chrm.prop.male.high1$strain == "MSM"]` in _molossinus^MSM^_ (Supplemental Figure 3). Following patterns in the genome-wide recombination rate, male _musculus^PWD^_ and male _molossinus^MSM^_ have 2CO proportions that are more similar to each other than to strains from their own subspecies (chi-square tests; _musculus^PWD^_ vs. *molossinus^MSM^*: p = 0.37 -- `r PWD.MSM.2CO.prop.test$p.value` ; _musculus^PWD^_ vs. _musculus^KAZ^_: p = 1.23^{-31} -- `r PWD.KAZ.2CO.prop.test$p.value`; _molossinus^MSM^_ vs. _molossinus^MOLF^_: p = 2.65^{-6} -- `r MSM.MOLF.2CO.prop.test$p.value` ). These results demonstrate that evolution of the genome-wide recombination rate reflects changes in crossover number across multiple bivalents.

<new 3/n>
To begin to localize evolution of genome-wide recombination rate to steps of the recombination pathway, we counted DMC1 foci in prophase spermatocytes as markers for double-strand breaks (DSBs). DMC1 foci were counted in a total of 76 -- `r sum( (DMC1table.main$ncells.x) )` early zygotene and 75 -- `r sum( (DMC1table.main$ncells.y) )` late zygotene spermatocytes from two high-recombination strains (*musculus^PWD^* and *molossinus^MSM^*) and three low-recombination strains (*musculus^KAZ^*, _domesticus^WSB^_, and _domesticus^G^_). High-recombination strains have significantly more DMC1 foci than low-recombination strains in early zygotene cells (t-test; p < 10^{-6} -- `r 1/1000000` ). In contrast, the two strain groups do not differ in DMC1 foci in late zygotene cells (t-test; p = 0.66 -- `r ttest.HighLow.Z.pval` ). Since DSBs are repaired as either COs or non-crossovers (NCOs), the ratio of MLH1 foci to DMC1 foci can be used to estimate the proportion of DSBs designated as COs. High-recombination and low-recombination strains do not differ in the MLH1/DMC1 ratio, whether DMC1 foci were counted in early zygotene cells or late zygotene cells (t-test; p > 0.05). These results raise the possibility that the evolution of genome-wide recombination rate is primarily determined by processes that precede the CO/NCO decision, at least in house mice.

<new 4/n>
Total SC length only partially differentiates high-recombination strains from low-recombination strains (Figure 5). Whereas high-recombination strains as a group have significantly greater total SC length than low-recombination strains (t-test; p = 0.01 -- `r Q2tot.sc_t.test_High.Low$p.value` ), separate tests within subspecies show that the two strain categories differ within *M. m. molossinus* (p = 0.02 -- `r Q2tot.sc_t.test_High.Low_Mol$p.value` ) but not within *M. m. musculus* (p = 0.4 -- `r Q2tot.sc_t.test_High.Low_Musc$p.value` ). Additionally, mouse means for the reduced (short and long) bivalent datasets do not differ between high-recombination and low-recombination strains (t-test; short: p = 0.88 -- `r Q2short.biv_t.test_High.Low$p.value`  ; long: p = 0.19 --  `r Q2long.biv_t.test_High.Low$p.value` ).

Write results for M4 and M5 here
total SC M4 results

Musc subspecies very ***, Mol *** (both ++)
G, LEW, 
MSM
SKIVE, KAZ (- -)

M5 (almost all strains are significant)
G, LEW, PWD, MSM
MOLF, 
SKIVE, KAZ, TOM, AST, CZECH

In a model with total SC length as the dependent variable (*M2?*), two subspecies effects are significant (*M. m. musculus* p = 2.33^{-6} -- `r Q2tot.sc_t.test_High.Low_Musc$p.value`, *M. m. molossinus*, p = 21^{-6} -- `r Q2tot.sc_t.test_High.Low_Mol$p.value`). In models with SC lengths of short and long bivalents as dependent variables, several subspecies and strain effects reach significance (p < 0.05), but they are not consistent across models. Collectively, these results reveal that evolution of SC length is not strongly associated with evolution of genome-wide recombination rate in house mice.

<new 5/n>
In summary, evolution of the genome-wide recombination rate in males is connected to double-strand break number and crossover interference, but not to SC length and crossover position (on single-crossover bivalents).

--------------------------

# DISCUSSION

<new 1/n>
By comparing recombination rates in females and males from the same diverse set of genetic backgrounds, we isolated sex as a primary factor in the evolution of this fundamental meiotic trait. Recombination rate differences are more pronounced in males than females. Because inter-strain divergence times are identical for the two sexes, this observation demonstrates that the genome-wide recombination rate evolves faster in males, at least in house mice. More generally, recombination rate divergence is decoupled in females and males. These disparities are remarkable given that recombination rates for the two sexes were measured in identical genomic backgrounds (other than the presence/absence of the Y chromosome). Our results provide the strongest evidence yet that the genome-wide recombination rate follows distinct evolutionary trajectories in males and females.

<new 2/n>
At the genetic level, the sex-specific evolution we documented indicates that some mutations responsible for divergence in recombination rate have dissimilar phenotypic effects in females and males. A subset of the genetic variants associated with genome-wide recombination rate within populations of humans [@Kong2004;, @Kong2008; @Kong2014; @halldorsson2019], Soay sheep [@johnston2016_soay], and cattle [@ma2015_cattle; @Shen2018_cattle] appear to show sex-specific properties, including opposite effects in females and males. Furthermore, inter-sexual correlations for recombination rate are weak in humans [@fledel2011] and Soay sheep [@johnston2016_soay]. Crosses between the strains we surveyed could be used to identify and characterize the genetic variants responsible for recombination rate evolution in house mice [@dumont2011; @Wang2017island; @wang2019_SC]. These variants could differentially affect females and males at any step in the recombination pathway. Although our DMC1 profiling was limited to males from a small number of strains (for practical reasons), our findings suggest that mutations that determine the number of double-strand breaks contribute to sex-specific evolution in the recombination rate. A study of two classical inbred strains and one wild-derived inbred strain of house mice also found a positive association between crossover number and double-strand break number in males [@baier2014].

<new 3/n>
Another implication of our results is that the connection between recombination rate and fitness differs between males and females. Little is known about whether and how natural selection shapes recombination rate in nature [@DapperPayseur2017; @Ritz2017].  @samuk2020 recently used a quantitative genetic test to conclude that an 8% difference in genome-wide recombination rate between females from two populations of *Drosophila pseudoobscura* was caused by natural selection. Applying similar strategies to species in which both sexes recombine, including house mice, would be a logical next step to understanding the sex-specific evolution of recombination rate.
